Bricking up backdoors in Intrusion Detection Systems

### 2019-09-13 paper/paper.tex
Several minor changes to paper

Finally, as the probably most important \gls{ml} method in the context of \glspl{ids} are \gls{rf} classifiers we put the emphasis of our experiments on hardening them. We propose a new pruning technique specifically for removing backdoors from trained \gls{rf} models.

We make all our code publicly available at \url{https://github.com/CN-TU/ids-backdoor}.

@@

\section{Experimental Setup} \label{sec:ml_approaches}
We performed our experiments with an \gls{rf} and a \gls{dl} model and intentionally added a backdoor to both. In particular, we used the following experimental setup.
\paragraph{Datasets}
%\todo{Max: I would shorten here.}
%Several datasets for the purpose of building and evaluating IDSs have been developed. However, as pointed out in \cite{gharib_evaluation_2016}, there are numerous requirements that have to be met for a dataset to provide realistic performance benchmarks in this context.

Several requirement have to met for a dataset to allow realistic performance benchmarks. % \cite{gharib_evaluation_2016}.
In this research, we use the \unsw{}~\cite{moustafa_unsw-nb15:_2015} and the CIC-IDS-2017~\cite{sharafaldin_toward_2018} datasets, which were developed by two independent institutions and are both freely available on the Internet. %, guaranteeing reproducibility of our results.

@@

The CIC-IDS-2017 dataset~\cite{sharafaldin_toward_2018} was created by the Canadian Institute of Cybersecurity to provide an alternative to existing datasets which are found to exhibit several shortcomings. The provided raw network captures contain more than 2.3 million flows, containing normal traffic and DoS, infiltration, web attacks, brute force and scanning attacks.

@@

%\subsubsection{\glspl{rf}}
In addition to \gls{dl}, we trained an \gls{rf} classifier to detect attacks in network traffic. We used the RandomForestClassifier implementation from scikit-learn for this task and used 100 estimators for training each \gls{rf}.

@@

We added a backdoor to our \gls{dl} and \gls{rf} models. % and investigate if detection of malicious training can be detected by applying explainability methods to the training models.
Hence,
%for this task
we
%first
had to find a pattern usable as a backdoor in feature vectors. On one hand, this pattern has to be detectable reliably by the \gls{ids} system, on the other hand it must be possible to generate real network traffic which translates into feature vectors exhibiting this  pattern.

In our case, we used the \gls{ttl} value contained in the \gls{ip} header of Internet traffic to signal the backdoor. The \gls{ttl} is used for mitigating problems due to routing loops in IP networks.
%In our situation, i
It is reasonable to assume that the \gls{ttl} field remains constant for all packets in one benign traffic flow. In particular, this assumption was perfectly reflected by the examined UNSW-NB15 dataset, as there were almost no traffic flows with a certain non-zero standard deviation of the \gls{ttl} value.

@@

\subsection{Performance Results}
\begin{table}[t]
\caption{Detection performance results.} \label{tab:performance_results}
\begin{tabular}{l r r r r} \toprule
& \multicolumn{2}{r}{\unsw{}} & \multicolumn{2}{r}{\cic{}} \\
& RF & DL & RF & DL \\ \midrule
%Accuracy	& 0.990 $\pm$ 0.000	& 0.989 $\pm$ 0.000	& 0.997 $\pm$ 0.000	& 0.998 $\pm$ 0.000	\\
%Precision	& 0.854 $\pm$ 0.001	& 0.845 $\pm$ 0.004	& 0.997 $\pm$ 0.000	& 0.999 $\pm$ 0.000	\\
%Recall	& 0.850 $\pm$ 0.003	& 0.829 $\pm$ 0.009	& 0.993 $\pm$ 0.000	& 0.992 $\pm$ 0.000	\\
%F1 score	& 0.852 $\pm$ 0.002	& 0.837 $\pm$ 0.003	& 0.995 $\pm$ 0.000	& 0.995 $\pm$ 0.000	\\
%Youden's J	& 0.845 $\pm$ 0.003	& 0.823 $\pm$ 0.009	& 0.992 $\pm$ 0.000	& 0.991 $\pm$ 0.000	\\
%Backdoor acc.	& 1.000 $\pm$ 0.000	& 0.998 $\pm$ 0.002	& 1.000 $\pm$ 0.000	& 1.000 $\pm$ 0.000	\\
Accuracy	& 0.990 & 0.989 & 0.997 & 0.998\\
Precision	& 0.854 & 0.845 & 0.997 & 0.999\\
Recall	& 0.850 & 0.829 & 0.993 & 0.992\\
F1 score	& 0.852 & 0.837 & 0.995 & 0.995\\
Youden's J	& 0.845 & 0.823 & 0.992 & 0.991\\
Backdoor acc.	& 1.000 & 0.998 & 1.000 & 1.000\\
\bottomrule
\end{tabular}
\end{table}
Table~\ref{tab:performance_results} shows performance results for both \gls{dl} and \gls{rf}, depicting both detection performance of normal samples as contained in the respective datasets and the efficacy of the backdoor. The models are thus able to detect the backdoor with high confidence while retaining high attack detection performance.
Our results are consistent with previous work like \cite{meghdouri_analysis_2018}.

@@

\subsection{Explainability Plots} \label{sec:plots}
%A major problem for the deployment of ML is the unpredictability and the lack of understanding of the decisions an ML models takes. For this reason, a
A number of methods have been proposed in the recent years aiming to visualize and explain a non-interpretable ML model's decisions.
Applied to the present problem, we can pick up ideas from \glspl{pdp} and \gls{ale} plots, not only for identifying backdoors in the \gls{mui}, but also for finding wrong decisions it would take due to flawed training data.

@@

\subsubsection{Accumulated Local Effects}
Due to feature dependence it is very likely that areas exist in the feature space which are very unlikely
% have a very low probability
to occur. Since a model is trained with real, observed data, the training set therefore does not include samples for these areas, which causes the model's predictions to become indeterminate for these areas. This poses a problem when considering these predictions for computing \glspl{pdp}.

In an attempt to overcome this problem, it is possible to only consider samples which are likely to occur for certain feature values, i.e. to consider the conditional distribution of remaining features, for computing explainability graphs. This results in the concept of \gls{ale} plots.

For the $i$th feature $X_i$, the \gls{ale} plot ALE$_i(w)$ can be defined differentially as

\begin{equation}
%\frac{d}{dw} \text{ALE}_i (w) = \mathbb E_{\boldsymbol X | X_i}\left(\frac{d}{dw} f(X_1,\ldots,X_{i-1},w,X_{i+1},\ldots X_n) \Big\vert X_i=w\right)
\frac{d}{dw} \text{ALE}_i (w) = \mathbb E_{\boldsymbol X | X_i}\left(\frac{\partial}{\partial X_i} f(\boldsymbol X) \, \Big \vert \, X_i=w\right)
\end{equation}
%\todo{I think that there are some minor problems with $y$ and $X_i$ in the formula.}

To combat ambiguity of this definition, we force ALE$_i(x)$ to have zero mean on the domain of $X_i$.
For empirical evaluation, we approximate the conditional distributions of $\boldsymbol X$ by averaging over samples for which $X_i \approx w$. In this paper, we used the 10  closest samples
%to the interval's center
for estimating the distribution.

@@

\subsubsection{Identifying Backdoors}
\begin{figure}[h]
\includegraphics[width=\columnwidth]{figures/pdpale2017nn_joint_2.pdf}
%\includegraphics[width=\columnwidth]{figures/pdpale2017nn.pdf}

%\includegraphics[width=\columnwidth]{figures/pdpale2017nn_zoom.pdf}
\caption{\gls{pdp} and \gls{ale} plots of the \gls{mlp} for the \cic{}. Full range of TTL values on top; only TTL values from 0 to 5 below.}
\label{fig:pdp_ttl}
\end{figure}
Backdoors can be identified by computing \gls{pdp} or \gls{ale} plots for the \gls{mui} and investigating if regions exist, for which the \gls{mui} behaves counter-intuitive. For our CIC-IDS-2017 \gls{dl} classifier, \autoref{fig:pdp_ttl} shows the \gls{pdp} for the \gls{ttl} value in forward direction. We also provide plots for the corresponding models which were trained without backdoor. The plots are not available in a real situation, but we provide them here for comparison.

As shown in \autoref{fig:pdp_ttl}, the \glspl{pdp} for \gls{dl} show a deep notch for certain low values of stdev(TTL). As discussed above, normal traffic is very unlikely to have deviating \gls{ttl} values for different packets. In contrast to \autoref{fig:pdp_ttl}, one would therefore expect this feature to have a negligible influence on the classification result. In our case, this notch thus points to the existence of the backdoor we embedded. For the \gls{rf} model the same pattern is observed but omitted for brevity.

However, inconsistent behaviour of the \gls{mui}, detected using \gls{pdp} or \gls{ale} plots, does not necessarily result from poisoning activity.  For example, \autoref{fig:ttlmean} shows the mean(TTL) feature in forward direction. The models show a clear dependence of the mean \gls{ttl} value of incoming packets, which is similarly counter-intuitive as for the feature discussed above. In our case, this behaviour results from the non-equal distribution of \gls{ttl} values of attack and non-attack traffic in both the \unsw{} and CIC-IDS-2017 datasets.

\begin{figure}[b]
\includegraphics[width=\columnwidth]{figures/ttlmean.pdf}
\caption{\gls{pdp} and \gls{ale} plots for mean(TTL) for the CIC-IDS-2017 \gls{rf} and \gls{dl} classifiers.}
\label{fig:ttlmean}
\end{figure}
Independent of the origin of such patterns, they might be exploited for masquerading attacks and thus are clearly unwanted. The use of \gls{pdp} and \gls{ale} plots therefore provides a convenient possibility for analyzing \gls{ml} models for such patterns.

@@

\begin{figure}[h]
\includegraphics[width=\columnwidth]{../prune_CAIA_backdoor_17/{prune_1.00_nn_0_bd}.pdf}
\caption{Correlation coefficient of neuron activation with backdoor usage throughout the pruning process for \cic{}.}
\label{fig:correlation}
\end{figure}

@@

\subsubsection{Fine-Tuning}
%\todo{Max: It said pruning before but should be tuning, right?}
%AH. I guess both should work
\begin{figure}[t]
%\includegraphics[width=\columnwidth]{figures/finetuning_2015.pdf}
\includegraphics[width=\columnwidth]{figures/finetuning_2017.pdf}
\caption{Fine-tuning of the \gls{mui} for \unsw{}.}
\label{fig:finetuning}
\end{figure}
In addition to pruning, we tried to use fine-tuning to remove the backdoor from the \gls{mlp} classifier.
The use of fine-tuning exclusively makes  sense if both the computational effort and the required training set size for fine-tuning are substantially lower than for the original training procedure.

>>>>

Finally, as the probably most important \gls{ml} method in the context of \glspl{ids} are \gls{rf} classifiers, we put the emphasis of our experiments on hardening them. We propose a new pruning technique specifically for removing backdoors from trained \gls{rf} models.

For reproducibility, we make all our code publicly available at \url{https://github.com/CN-TU/ids-backdoor}.

@@

\section{Experimental Setup} \label{sec:ml_approaches}
We performed our experiments with an \gls{rf} and a \gls{mlp} model and intentionally added a backdoor to both. In particular, we used the following experimental setup.
\paragraph{Datasets}
%\todo{Max: I would shorten here.}
%Several datasets for the purpose of building and evaluating IDSs have been developed. However, as pointed out in \cite{gharib_evaluation_2016}, there are numerous requirements that have to be met for a dataset to provide realistic performance benchmarks in this context.

Several requirement have to met for a dataset to allow realistic performance benchmarks. % \cite{gharib_evaluation_2016}.
In this research, we use the \unsw{}~\cite{moustafa_unsw-nb15:_2015} and the \cic{}~\cite{sharafaldin_toward_2018} datasets, which were developed by two independent institutions and are both freely available on the Internet. %, guaranteeing reproducibility of our results.

@@

The \cic{} dataset~\cite{sharafaldin_toward_2018} was created by the Canadian Institute of Cybersecurity to provide an alternative to existing datasets which are found to exhibit several shortcomings. The provided raw network captures contain more than 2.3 million flows, containing normal traffic and DoS, infiltration, web attacks, brute force and scanning attacks.

@@

%\subsubsection{\glspl{rf}}
In addition to the \gls{mlp}, we trained an \gls{rf} classifier to detect attacks in network traffic. We used the RandomForestClassifier implementation from scikit-learn for this task and used 100 estimators for training each \gls{rf}.

@@

We added a backdoor to our \gls{mlp} and \gls{rf} models. % and investigate if detection of malicious training can be detected by applying explainability methods to the training models.
Hence,
%for this task
we
%first
had to find a pattern usable as a backdoor in feature vectors. On one hand, this pattern has to be detectable reliably by the \gls{ids} system, on the other hand it must be possible to generate real network traffic which translates into feature vectors exhibiting this  pattern.

In our case, we used the \gls{ttl} value contained in the \gls{ip} header of Internet traffic
as channel for signaling the backdoor.
%to signal the backdoor.
The \gls{ttl} is used for mitigating problems due to routing loops in IP networks.
%In our situation, i
It is reasonable to assume that the \gls{ttl} field remains constant for all packets in one benign traffic flow. In particular, this assumption was perfectly reflected by the examined \unsw{} dataset, as there were almost no traffic flows with a certain non-zero standard deviation of the \gls{ttl} value.

@@

\subsection{Performance Results}
\begin{table}[t]
\caption{Detection performance results.} \label{tab:performance_results}
\begin{tabular}{l r r r r} \toprule
& \multicolumn{2}{r}{\unsw{}} & \multicolumn{2}{r}{\cic{}} \\
& RF & DL & RF & DL \\ \midrule
%Accuracy	& 0.990 $\pm$ 0.000	& 0.989 $\pm$ 0.000	& 0.997 $\pm$ 0.000	& 0.998 $\pm$ 0.000	\\
%Precision	& 0.854 $\pm$ 0.001	& 0.845 $\pm$ 0.004	& 0.997 $\pm$ 0.000	& 0.999 $\pm$ 0.000	\\
%Recall	& 0.850 $\pm$ 0.003	& 0.829 $\pm$ 0.009	& 0.993 $\pm$ 0.000	& 0.992 $\pm$ 0.000	\\
%F1 score	& 0.852 $\pm$ 0.002	& 0.837 $\pm$ 0.003	& 0.995 $\pm$ 0.000	& 0.995 $\pm$ 0.000	\\
%Youden's J	& 0.845 $\pm$ 0.003	& 0.823 $\pm$ 0.009	& 0.992 $\pm$ 0.000	& 0.991 $\pm$ 0.000	\\
%Backdoor acc.	& 1.000 $\pm$ 0.000	& 0.998 $\pm$ 0.002	& 1.000 $\pm$ 0.000	& 1.000 $\pm$ 0.000	\\
Accuracy	& 0.990 & 0.989 & 0.997 & 0.998\\
Precision	& 0.854 & 0.845 & 0.997 & 0.999\\
Recall	& 0.850 & 0.829 & 0.993 & 0.992\\
F1 score	& 0.852 & 0.837 & 0.995 & 0.995\\
Youden's J	& 0.845 & 0.823 & 0.992 & 0.991\\
Backdoor acc.	& 1.000 & 0.998 & 1.000 & 1.000\\
\bottomrule
\end{tabular}
\end{table}
Table~\ref{tab:performance_results} shows performance results for both the \gls{mlp} and \gls{rf}, depicting both detection performance of normal samples as contained in the respective datasets and the efficacy of the backdoor. The models are thus able to detect the backdoor with high confidence while retaining high attack detection performance.
Our results are consistent with previous work like \cite{meghdouri_analysis_2018}.

@@

\subsection{Explainability Plots} \label{sec:plots}
%A major problem for the deployment of ML is the unpredictability and the lack of understanding of the decisions an ML models takes. For this reason, a
A number of methods have been proposed recently aiming to visualize and explain a non-interpretable ML model's decisions.
Applied to the present problem, we can pick up ideas from \glspl{pdp} and \gls{ale} plots, not only for identifying backdoors in the \gls{mui}, but also for finding wrong decisions it would take due to flawed training data.

@@

\subsubsection{Accumulated Local Effects}
Due to feature dependence it is very likely that areas exist in the feature space which are very unlikely
% have a very low probability
to occur. Since a model is trained with real, observed data, the training set therefore does not include samples for these areas, which causes the model's predictions to become indeterminate for these areas, %. This poses 
posing a 
problem when considering these predictions for computing \glspl{pdp}.

In an attempt to overcome this problem, it is possible to only consider samples which are likely to occur for certain feature values, i.e. to consider the conditional distribution of remaining features, for computing explainability graphs.

\gls{ale} plots~\cite{apley_visualizing_2016} make use of this idea.
For the $i$th feature $X_i$, the \gls{ale} plot ALE$_i(w)$ can be defined differentially as
\begin{equation}
%\frac{d}{dw} \text{ALE}_i (w) = \mathbb E_{\boldsymbol X | X_i}\left(\frac{d}{dw} f(X_1,\ldots,X_{i-1},w,X_{i+1},\ldots X_n) \Big\vert X_i=w\right)
\frac{d}{dw} \text{ALE}_i (w) = \mathbb E_{\boldsymbol X | X_i}\left(\frac{\partial}{\partial X_i} f(\boldsymbol X) \, \Big \vert \, X_i=w\right) .
\end{equation}
%\todo{I think that there are some minor problems with $y$ and $X_i$ in the formula.}

\begin{figure}[b!]
\includegraphics[width=\columnwidth]{figures/pdpale2017nn_joint_2.pdf}
%\includegraphics[width=\columnwidth]{figures/pdpale2017nn.pdf}

%\includegraphics[width=\columnwidth]{figures/pdpale2017nn_zoom.pdf}
\caption{\gls{pdp} and \gls{ale} plots of the \gls{mlp} for the \cic{}. Full range of TTL values on top; only TTL values from 0 to 5 below.}
\label{fig:pdp_ttl}
\end{figure}

To combat ambiguity of this definition, we force ALE$_i(w)$ to have zero mean on the domain of $X_i$.
For empirical evaluation, we approximate the conditional distributions of $\boldsymbol X$ by averaging over samples for which $X_i \approx w$. In this paper, we used the 10  closest samples
%to the interval's center
for estimating the distributions.

@@

\subsubsection{Identifying Backdoors}

Backdoors can be identified by computing \gls{pdp} or \gls{ale} plots for the \gls{mui} and investigating if regions exist, for which the \gls{mui} behaves counter-intuitive. For our \cic{} \gls{mlp} classifier, \autoref{fig:pdp_ttl} shows the \gls{pdp} for the \gls{ttl} value in forward direction. We also provide plots for the corresponding models which were trained without backdoor. The plots are not available in a real situation, but we provide them here for comparison.

As shown in \autoref{fig:pdp_ttl}, the \glspl{pdp} for the \gls{mlp} show a deep notch for certain low values of stdev(TTL). As discussed above, normal traffic is very unlikely to have deviating \gls{ttl} values for different packets. In contrast to \autoref{fig:pdp_ttl}, one would therefore expect this feature to have a negligible influence on the classification result. In our case, this notch thus points to the existence of the backdoor we embedded. For the \gls{rf} model the same pattern is observed but omitted for brevity.

However, inconsistent behaviour of the \gls{mui}, detected using \gls{pdp} or \gls{ale} plots, does not necessarily result from poisoning activity.  For example, \autoref{fig:ttlmean} shows the mean(TTL) feature in forward direction. The models show a clear dependence of the mean \gls{ttl} value of incoming packets, which is similarly counter-intuitive as for the feature discussed above. In our case, this behaviour results from the non-equal distribution of \gls{ttl} values of attack and non-attack traffic in both the \unsw{} and \cic{} datasets.

Independent of the origin of such patterns, they might be exploited for masquerading attacks and thus are clearly unwanted. The use of \gls{pdp} and \gls{ale} plots therefore provides a convenient possibility for analyzing \gls{ml} models for such patterns.

@@

\begin{figure}[t]
\includegraphics[width=\columnwidth]{figures/ttlmean.pdf}
\caption{\gls{pdp} and \gls{ale} plots for mean(TTL) for the \cic{} \gls{rf} and \gls{mlp} classifiers.}
\label{fig:ttlmean}
\end{figure}

\begin{figure}[b]
\includegraphics[width=\columnwidth]{../prune_CAIA_backdoor_17/{prune_1.00_nn_0_bd}.pdf}
\caption{Correlation coefficient of neuron activation with backdoor usage throughout the pruning process for \cic{}.}
\label{fig:correlation}
\end{figure}

@@

\subsubsection{Fine-Tuning}
%\todo{Max: It said pruning before but should be tuning, right?}
%AH. I guess both should work
\begin{figure}[b]
%\includegraphics[width=\columnwidth]{figures/finetuning_2015.pdf}
\includegraphics[width=\columnwidth]{figures/finetuning_2017.pdf}
\caption{Fine-tuning of the \gls{mui} for \unsw{}.}
\label{fig:finetuning}
\end{figure}
In addition to pruning, we tried to use fine-tuning to remove the backdoor from the \gls{mlp} classifier.
The use of fine-tuning exclusively makes  sense if both the computational effort and the required training set size for fine-tuning are substantially lower than for the original training procedure.
